\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2020-04-22 analysis/paper/paper.tex
last edits for draft
Fixed code formatting, typos, plot colors.

we have a total of 9104 observations and 30 variables, two of which are
SUPPORT day 3 physiology score (sps) and APACHE III day 3 physiology
score (aps). These scores were developed for a period of 180 days, which
makes our comparison over 5.56 years exploratory. Therefore, we will
assess these two models before using either as our baseline, by
comparing them to a full model using all variables. Also, our single
imputation introduces bias in the data. As the goal here is purely
demonstrative, this bias will be consistent with any of the following
models. As such, their comparison is still valid, though any
interpretation requires an awareness of said bias. A description of used
variables can be found in Table \ref{tab:support1} and a breakdown of
each categorical variable in Table \ref{tab:support2}. The original data
was obtained from the Department of Biostatistics at Vanderbilt
University \citep{harrell_2020}.

@@

Before generating an absolute risk curve using regularized logistic
regression, we will explore three options as a baseline comparison,
beyond a Kaplan-meier curve. The original study compared these scores
over a span of 180 days, which may not be comparable over the full 5.56
years. Therefore, We will examine three cox regression scenarios: sps
only, aps only and all covariates excluding sps and aps. Next, we will
compare each modeling using the Brier score \citep{graf1999ass}. The
Brier score is able to demonstrate both discrimination and calibration
in a single model \citep{graf1999ass} and provide a comparable metric
between semi-parametric and parametric methods in the survival context.
For all the following analyses in this case study, we will use 80\% of
the observations to train our models, while 20\% will be used to
generate absolute risk curves, all of which will be averaged to
demonstrate each model's predictive ability. Over 5.56 years,
\ref{fig:baselines} demonstrates that all the cox models have similar,
if not identical average predictions. Figure \ref{fig:brier1} examines
the brier scores as they change over time. As the full model has the
lowest brier score on overall, it will be used in the following absolute
risk comparison to regularized cox, regularized logistic regression and
the unadjusted Kaplan-Meier curve.

@@

To get the absolute risk following regularized cox regression, we had to
work around limitations in existing packages. To make use of the
\pkg{survival} package, we first get our hazard coefficients using
\pkg{glmnet}'s coxnet implementation for lasso penalization
\citep{regpathcox}. Next, we fit a cox model using the survival package
with the selected variables only, from our previous fit. Finally, we
replaced the coefficients in the survival object with those from our
glmnet fit, allowing us to make use of the absolute risk tools available
in the \pkg{survival} package.

@@

\begin{CodeInput}
R> # hazard for cox using glmnet
R> coxNet <- glmnet::cv.glmnet(x = x, y = y, family = "cox", alpha = 1, standardize = TRUE)
R> #getting all non-zero covariates
R> nonzero_covariate_cox <- predict(coxNet, type = "nonzero", s = "lambda.1se")
R> #keeping coefficients for later
R> nonzero_coef_cox <- coef(coxNet, s = "lambda.1se")
R> # cutting dataset to only include nonzero covariates
R> cleanCoxData <- as.data.frame(cbind(y, x[, nonzero_covariate_cox]))
R> #create a skeleton object
R> coxNet <- survival::coxph(Surv(time = d.time, event = death) ~ ., data = cleanCoxData,x=TRUE)
R> #replace coefficients with regularized ones
R> coxNet$coefficients <- nonzero_coef_cox@x
\end{CodeInput}
\end{CodeChunk}

For our penalized logistic regression model, we opted for the natural
log of time, which assumes a Weibull distribution for our event. As
casebase makes use of the glmnet package, we interact with the
fitsmoothhazard.fit function using a matrix interface, where y contains
the time and event variables and x contains all other variables we would
like to include in the model. For the SUPPORT dataset, all factor
variables were converted into dummy variables. Here, we used lasso
penalization by setting alpha to 1, leading to our smooth hazard
function.

@@

\begin{CodeInput}
R> #regularized logistic regression to estimate hazard
R> cbFull <- casebase::fitSmoothHazard.fit(x, y, family = "glmnet",
R>                       time = "d.time", event = "death",
R>                       formula_time = ~ log(d.time), alpha = 1,
R>                       ratio = 100, standardize = TRUE, penalty.factor=c(0,rep(1,ncol(x))))
R> 
R> # Estimating the absolute risk curve using the newData parameter.
R> abCbFull <- casebase::absoluteRisk(cbFull, time = times, newdata = newData, s = "lambda.1se", method = c("numerical"))
R> 
\end{CodeInput}
\end{CodeChunk}

\begin{CodeInput}
R> 
R> cbFull <- casebase::fitSmoothHazard.fit(x, y, family = "glmnet", time = "d.time", event = "death", formula_time = ~ log(d.time), alpha = 1, ratio = 10, standardize = TRUE,penalty.factor=c(0,rep(1,ncol(x))))
\end{CodeInput}
\end{CodeChunk}

\begin{CodeInput}
R> # Estimating the absolute risk curve using the newData parameter.
R> abCbFull <- casebase::absoluteRisk(cbFull, time = times, newdata = newData, s = "lambda.1se", method = c("numerical"))
\end{CodeInput}
\end{CodeChunk}

@@

\caption{\label{fig:cs3lolliPlot} A lollipop plot demonstrating the estimated coefficients between the following three models: Cox (green), penalized Cox (red), penalized logisitic (blue). If a coefficient is estimated to be 0, it is removed from the plot}\label{fig:coefplots}
\end{figure}
\end{CodeChunk}

@@

Our first comparison plot examines the covariates that were selected
between all three hazard functions. Note that Cox regression without
penalization selected more covariates than the other two, but only the
intersection of cofficients with values greater than 0 for all models
are displayed in \ref{fig:cs3lolliPlot}. A visible tendency is for Cox
regression to have larger estimated coefficients. This is expected, as
there are no restrictions on the size of coefficients. The smallest
coefficient estimates are found in penalized logistic model, aside from
cancer state (cano). These values do not inform us of model performance.
To do so, we take our hazard functions and use them to predict the
absolute risk of each individual in our test set, over time. The
probabilites over time for each individual are averaged, resulting in
the absolute risk curves in figure \ref{fig:cs3Final}. There is
separation between the curves, but it is minimal. To better demonstrate
the differences between the models, we compare brier scores over time.
\ref{fig:cs3FinalBrier} the Brier score is lower in our penalized
logistic regression model, compared to the other models. This
difference, however, is minimal. We can say with relative certainty that
any of the hazard models outperform an unadjusted curve, while penalized
logistic regression performs the best on this testing dataset.

@@

This report was generated on 2020-04-16 12:08:46 using the following
computational environment and dependencies:

@@

\begin{CodeOutput}
#> Local:    master C:/Users/jesse/Documents/gits/cbpaper
#> Remote:   master @ origin (https://github.com/Jesse-Islam/cbpaper.git)
#> Head:     [7755d8f] 2020-04-09: Uses development version of Jesse-Islam/riskRegression
\end{CodeOutput}
\end{CodeChunk}

>>>>

we have a total of 9104 observations and 30 variables, two of which are
SUPPORT day 3 physiology score (sps) and APACHE III day 3 physiology
score (aps). These scores were developed for a period of 180 days, which
makes our comparison over 5.56 years exploratory. Therefore, we will
assess these two models before using either as our baseline, by
comparing them to a full model using all variables. Also, our single
imputation introduces bias in the data. As the goal here is purely
demonstrative, this bias will be consistent with any of the following
models. As such, their comparison is still valid, though any
interpretation requires an awareness of said bias. A description of
covariates and response variables can be found in Table
\ref{tab:support1} and a breakdown of each categorical variable in Table
\ref{tab:support2}. The original data was obtained from the Department
of Biostatistics at Vanderbilt University \citep{harrell_2020}.

@@

Before generating an absolute risk curve using regularized logistic
regression, we will explore three options as a baseline comparison. The
original study compared these scores over a span of 180 days, which may
not be comparable over the full 5.56 years. Therefore, We will examine
three cox regression scenarios: sps only, aps only and all covariates
excluding sps and aps. Next, we will compare each modeling using the
Brier score \citep{graf1999ass}. The Brier score can demonstrate both
discrimination and calibration in a single model \citep{graf1999ass} and
provide a comparable metric between semi-parametric and parametric
methods in the survival context. For all the following analyses in this
case study, we will use 80\% of the observations to train our models,
while 20\% will be used to generate absolute risk curves, all of which
will be averaged to demonstrate each model's predictive ability. Over
5.56 years, \ref{fig:baselines} demonstrates that all the cox models
have similar, if not identical average predictions. Figure
\ref{fig:brier1} examines the brier scores as they change over time. As
the full model has the lowest brier score on overall, it will be used in
the following absolute risk comparison to regularized cox, regularized
logistic regression and the unadjusted Kaplan-Meier curve.

@@

To get the absolute risk following regularized cox regression, we had to
work around limitations in existing packages. To make use of the
\pkg{survival} package, we first estimate the hazard coefficients using
thecoxnet implementation for lasso penalization in \pkg{glmnet}
\citep{regpathcox}. Next, we fit a cox model using the survival package
with the selected variables only, from our regularized fit. Finally, we
replaced the coefficients in the survival object with those from our
glmnet fit, allowing us to make use of the absolute risk tools available
in the \pkg{survival} package.

@@

\begin{CodeInput}
R> # hazard for cox using glmnet
R> coxNet <- glmnet::cv.glmnet(
R>   x = x, y = y, family = "cox", alpha = 1,
R>   standardize = TRUE
R> )
R> # getting all non-zero covariates
R> nonzero_covariate_cox <- predict(coxNet,
R>   type = "nonzero",
R>   s = "lambda.1se"
R> )
R> # keeping coefficients for later
R> nonzero_coef_cox <- coef(coxNet, s = "lambda.1se")
R> # cutting dataset to only include nonzero covariates
R> cleanCoxData <- as.data.frame(cbind(y, x[, nonzero_covariate_cox]))
R> # create a skeleton object
R> coxNet <- survival::coxph(Surv(time = d.time, event = death) ~ .,
R>   data = cleanCoxData, x = TRUE
R> )
R> # replace coefficients with regularized ones
R> coxNet$coefficients <- nonzero_coef_cox@x
\end{CodeInput}
\end{CodeChunk}

For our penalized logistic regression model, we opted for the natural
log of time, which assumes a Weibull distribution for our event. As
casebase makes use of the glmnet package, we interact with the
fitsmoothhazard.fit function using a matrix interface, where \emph{y}
contains the time and event variables and \emph{x} contains all
covariates we would like to include in the model. For the SUPPORT
dataset, all factor variables were converted into dummy variables. Here,
we used lasso penalization by setting alpha to 1, leading to our smooth
hazard function.

@@

\begin{CodeInput}
R> # regularized logistic regression to estimate hazard
R> cbFull <- casebase::fitSmoothHazard.fit(x, y,
R>   family = "glmnet",
R>   time = "d.time", event = "death",
R>   formula_time = ~ log(d.time), alpha = 1,
R>   ratio = 100, standardize = TRUE,
R>   penalty.factor = c(0, rep(1, ncol(x)))
R> )
R> 
R> # Estimating the absolute risk curve using the newData parameter.
R> abCbFull <- casebase::absoluteRisk(cbFull,
R>   time = times,
R>   newdata = newData,
R>   s = "lambda.1se",
R>   method = c("numerical")
R> )
\end{CodeInput}
\end{CodeChunk}

@@

\caption{\label{fig:cs3lolliPlot} A lollipop plot demonstrating the estimated coefficients between the following three models: Cox (green), penalized Cox (red), penalized logisitic (blue). If a coefficient is estimated to be 0, in any model, it is removed from the plot for simplicity.}\label{fig:coefplots}
\end{figure}
\end{CodeChunk}

@@

Our first comparison plot examines the covariates that were selected
between all three hazard functions. Note that Cox regression without
penalization selected more covariates than the other two, but only the
intersection of coefficients with values greater than 0 are displayed in
\ref{fig:cs3lolliPlot}. A visible tendency is for Cox regression to have
larger estimated coefficients. This is expected, as there are no
restrictions on the size of coefficients. The smallest coefficient
estimates are found in penalized logistic model, aside from cancer state
(cano). These values do not inform us of model performance. To do so, we
take our hazard functions and use them to predict the absolute risk of
everyone in our test set, over time. The probabilities over time for
each individual are averaged, resulting in the absolute risk curves in
figure \ref{fig:cs3Final}. There is separation between the curves, but
it is minimal. To better demonstrate the differences between the models,
we compare brier scores over time. \ref{fig:cs3FinalBrier} The Brier
score is lower in our penalized logistic regression model, compared to
the other models. This difference, however, is minimal. We can say with
relative certainty that any of the hazard models outperform an
unadjusted curve, while penalized logistic regression performs the best
on this testing dataset.

@@

This report was generated on 2020-04-22 10:48:55 using the following
computational environment and dependencies:

@@

\begin{CodeOutput}
#> Local:    master C:/Users/jesse/Documents/gits/cbpaper
#> Remote:   master @ origin (https://github.com/Jesse-Islam/cbpaper.git)
#> Head:     [d079224] 2020-04-16: Write up and plots fixed. plot descriptions fixed
\end{CodeOutput}
\end{CodeChunk}
